%%
%% Cow Pi Datasheet (c) 2021-23 Christopher A. Bohn
%%
%% Licensed under the Apache License, Version 2.0 (the "License");
%% you may not use this file except in compliance with the License.
%% You may obtain a copy of the License at
%%     http://www.apache.org/licenses/LICENSE-2.0
%% Unless required by applicable law or agreed to in writing, software
%% distributed under the License is distributed on an "AS IS" BASIS,
%% WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
%% See the License for the specific language governing permissions and
%% limitations under the License.
%%

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{arydshln}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{etoolbox}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[procnames]{listings}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{subfig}
\usepackage{textcomp}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric}
\usepackage{xcolor}
\usepackage{multirow}

\lstset{language=C, tabsize=4, upquote=true, basicstyle=\ttfamily, numberstyle=\color{gray}, numbers=left, showspaces=false, showstringspaces=false}
\newcommand{\function}[1]{\textbf{\lstinline{#1}}}
\newcommand{\functionitem}[3]{\item \hypertarget{function:#2}{\lstinline{#1} \function{#2} \lstinline{(#3)}}\addcontentsline{toc}{subsubsection}{\function{#2}}}
\hypersetup{colorlinks=true}

\input{macros}

\begin{document}
    \pagestyle{fancy}
    \fancyhf{}
    \lhead{Cow Pi \cowpiversion\ Development Board}
    \rhead{CowPi v\softwareversion\ Library}
    \cfoot{\thepage}
    \title{Cow Pi, \hardwareversion\footnote{The Cow Pi hardware design, library, example code, and documentation \\ \copyright 2021--23, Christopher A. Bohn\tiny{, and licensed under the Apache~2.0 License. See \url{https://github.com/DocBohn/CowPi/} for details.}}}
    \author{Datasheet}
    \date{}
    \maketitle

    The \hardwareversion\ Cow Pi is a member of the Cow Pi product line from Eclectic Electronics\footnote{
        Eclectic Electronics is a joint subsidiary of The Pleistocene Petting Zoo, a Limited-Reality Corporation, and of Babbage's Analytic Engines, a Frankly-Figmentary Partnership. \\
        \tiny{(These are fictional companies imagined by Christopher Bohn for course assignments. You didn't think they really existed, did you? \\ Christopher Bohn, however, really does exist and can be contacted at \href{mailto:bohn@unl.edu}{bohn@unl.edu}.)}
    }
    and is guaranteed to bring you minutes of fun.
    Whether you need a 4-function calculator to count your terror bird eggs before they hatch, an electronic combination lock to secure your wildly-dangerous experimental robots, or a motion alarm to protect yourself from zombie gourds, using the Cow Pi development board is your best option.

    This document describes the use of the Cow Pi \cowpiversion\ development board with the CowPi v\softwareversion\ library.

    \tableofcontents


    \section{Hardware Overview} \input{hardware}

    \section{Library Overview} \input{software}

    \section{Example Programs} \input{examples}

    \section{Input/Output Register Descriptions} \label{sec:IOregisterDescriptions} \input{atmega328p-registers}

    \section{Interrupts} \label{sec:Interrupts} \input{atmega328p-interrupts}

    \section{Timers} \label{sec:Timers} \input{atmega328p-timers}

    \section{Expansion Options}

    This section reserved for future use.

    \bibliographystyle{plain}\bibliography{datasheet}

\end{document}

